Brief Announcement: An Improved Lower Bound for Dynamic Graph Partitioning

### 2020-01-29 paper/paper.tex
Sections Model and Lower bound

Formally, the online \emph{Balanced Re-Partitioning} problem (BRP) is defined as
follows. There is a set of $n$ nodes, initially distributed arbitrarily
across $\ell$~clusters, each of size~$k$. We call two nodes~$u,v\in V$
\emph{collocated} if they are in the same cluster.

An input to the problem is a sequence of communication requests $\sigma =
(u_1,v_1),$ $(u_2,v_2),$ $(u_3,v_3), \ldots$, where pair $(u_t,v_t)$ means that
the nodes $u_t,v_t$ exchange a fixed amount of data.  At any time~$t$, an online algorithm needs to serve the~communication
request~$(u_t,v_t)$. Right before serving the request, the online algorithm
can repartition the nodes into new clusters. We assume that
a~communication request between two collocated nodes costs 0. The cost of a~communication request between two nodes located in different clusters is
normalized to~1, and the cost of migrating a node from one cluster to another
is~$\alpha \geq 1$, where $\alpha$ is a parameter (an~integer). For any
algorithm $ALG$, we denote its total cost (consisting of communication plus
migration costs) on sequence $\sigma$ by $ALG(\sigma)$.

\section{Lower bound $\Omega(k\cdot \ell)$ for Competitive Ratio of Any Deterministic Algorithm}

In this section we consider the Online Balanced Partition problem.
We show a lower bound of $\Omega(k \cdot \ell)$ that uses an input that has a perfect partition.
Therefore, this result is a lower bound for the general Online Balanced Partition problem, as well as a lower bound for Online Balanced Partition with perfect partition.
The latter complements an upper bound $O(k \cdot \ell)$ for Online Balanced Partition with perfect partition that we present in Section \ref{sec:upperbound}.

\begin{theorem}
  The competitive ratio of any deterministic algorithm for Online Balanced Partition is $\Omega(k\cdot \ell)$.
\end{theorem}

\begin{proof}
We construct an instance of the problem with $\ell$ clusters 
$\set{ S_1, S_2,\dots , S_{\ell}}, |S_i|  = k$.
Let $I(C)$ denote the cluster where nodes in a component $C$ are located initially.
Consider any online algorithm ALG.
We construct the input sequence of requests for ALG as follows.
First,
we issue $k-2$ (internal) requests so that ALG form a component of $k-1$
nodes on clusters $S_1$.
Note that this does not cost ALG.
Let $x_0$  be the only single nodes left on $S_1$ and  $y_0 \in S_2$ be any single node on $S_2$.
Next,
we issue the request between $x_0$ and $y_0$.
Since this is an external request,
ALG joins them into one component and collocates them in some cluster other than $S_1$.
For this,
ALG moves to a new configuration that replaces $x_0$ and $y_0$ with two other single nodes $x_1$ and $y_1$ respectively.
Next,
we issue a request between $x_1$ and the largest component $C$ s.t.~$I(C) = I(x_1)$.
ALG must collocate $x_1$ and $C$ on some cluster other than $S_1$ and
consequently replaces $x_1$.
We repeat issuing requests between the single node $x_i$ on $S_1$ and the largest component $C'$ s.t.~$I(C')=I(x_i)$,
 until there are only two single nodes left that  originate from the same cluster.
I.e.,
two single nodes $x^*, y^*,I(x^*) = I(y^*)$
that constitute the last pair of such nodes.
At this point there are at most $\ell+1$ single nodes left
since there must be at least two nodes with the same initial cluster in any subset of $\ell+1$
nodes.
Given this sequence of requests,
the optimal strategy is to migrate $\set{x_0,y_0}$ to the cluster $I(x^*)$ by
 swapping $x_0$ and $y_0$ with $x^*$ and $y^*$ respectively.
Hence,
OPT pays $4$ for node migrations and
ALG pays at least one for each node in the sequence $X := x_0, x_1,\dots$.
We exclude at most $(k-1) + ( \ell+1)$ nodes out of $k.\ell$ nodes,
therefore $|X| \geq k.\ell - k - \ell \in \Omega(k\cdot\ell)$.
\end{proof}

\section{Perfect Partition Learner Algorithm}

@@

\begin{algorithm}
    \renewcommand{\algorithmicrequire}{\textbf{Input:}}
    \renewcommand{\algorithmicensure}{\textbf{Output:}}
    \begin{algorithmic}[1]
        \Require 
        $k, \ell$,
        initial partition $P_I$,
        sequence of  requests $\sigma_1, \dots, \sigma_N$ 
        \Ensure A final partition $P_F$ 
        \State For each node $v$ create a singleton component $C_v$ and add it to $\mathcal{C}$
        \State{$P_0 := P_I$}
         \label{line:initcomponents}
        \For {each  request $\sigma_t=\{u,v\}, 1 \leq t \leq N$}
        \State Let $C_1 \ni u$ and $C_2 \ni v$ be the container components
        \If{$C_1 \neq C_2$}
        \State Unite the two components into a single component $C'$ and
        $\mathcal{C} = (\mathcal{C}\setminus\set{C_1, C_2}) \cup ~\set{C'}$ \label{line:mergecomponents}
%        \If{$\mathit{cluster}(C_1, P_{t-1}) \neq \mathit{cluster}(C_2, P_{t-1})$}
        \If{$C_1$ and $C_2$ are not in the same cluster w.r.t.~$P_{t-1}$}
        \State $P_{t} = \mathit{re\mhyphen partition}(P_{t-1}, P_I, \mathcal{C}$) \label{line:rebalance} 
        \Comment{move to $P$ closest to $P_I$}
        \EndIf
        \EndIf
        \EndFor
    \end{algorithmic}
    \caption{Perfect Partition Learner (PPL)}
    \label{alg:ppl}
      \end{algorithm}

@@

Let \emph{Greedy} be the algorithm that on each inter-cluster request
 moves to a perfect partition closest to the current one.
 That is,
 a partition reachable with the least number of node swaps from the current partition.

@@

\begin{lemma} \label{lemma:k=3}
  The re-balancing cost of Greedy for $k=3$ is $O(1)$.
  %\maciek{Greedy = find the cheapest (closest to the current) feasible partition}
  \label{rebalancing-cost}
\end{lemma}

@@

\begin{theorem}
	Algorithm \emph{Greedy} is $O(\ell)$-competitive when $k=3$.
\end{theorem}

\begin{proof}
	The adversary issues at most $3\ell$ requests, as otherwise the perfect partition would not exist.
	Each request triggers at most one re-balancing in \emph{Greedy}
	which costs $O(1)$ by Theorem~\ref{rebalancing-cost}.
	Then,
	the total cost of any algorithm is $O(\ell)$.
\end{proof}

We leverage the simple algorithm \emph{Greedy} for $k=3$
towards an algorithm for the general problem
where the communication graph is not restricted.
Note that in the general model,
an optimal partition is not a necessarily a perfect partition.
We divide the sequence of requests into phases.
A phase terminates when there is no perfect partition of components.
We apply requests to \emph{Greedy} as before. 
Once the current phase terminates,
we reset all components by removing (forgetting) all revealed edges,
and then we proceed to the next phase.

\begin{theorem} \label{cor:k=3}
	There exists $O(\ell)$-competitive algorithm for the general online problem when $k=3$. 
\end{theorem}
\begin{proof}
	Consider any phase and assume it terminates upon arrival of some request $r$.
	By Lemma  \ref{lemma:k=3},
	\emph{Greedy} pays $O(\ell)$ during the phase.
	If OPT stays in the same configuration throughout the phase,
	regardless of its partition at the beginning of the phase,        
	it incurs remote communication cost for at least one request $r'$ (possibly $r'=r$) by the end of the phase.
	Else,
	OPT incurs the cost of moving to new partitions.
	In any case,
	OPT pays at least one during the phase and 
	the competitive ratio is $O(\ell)$.
\end{proof}

>>>>

Formally, the online \emph{Balanced Re-Partitioning} problem (OBRP) is defined as
follows. There is a set of $n$ nodes,
initially arbitrarily partitioned into $\ell$~clusters,
each of size~$k$.
Nodes in~$C \subset V$ are \emph{collocated} by the partition $P$
if $P$ assigns them to the same cluster.
I.e.,
$\forall u,v \in C: \mathit{cluster} (u,P) = \mathit{cluster} (v,P)$.
The input to the problem is a sequence of communication requests
$\sigma = (u_1,v_1),$ $(u_2,v_2),$ $(u_3,v_3), \ldots$,
where a pair $(u_t,v_t)$ indicates that nodes $u_t,v_t$ exchange a fixed amount of data.
An online algorithm needs to serve the communication request~$(u_t,v_t)$
as soon as it arrives.
Before serving the request,
the online algorithm
can re-partition,
i.e.,
it may move some of the nodes into clusters different than their current assigned clusters.
We assume that
a~communication request between two collocated nodes costs 0,
the cost of a~communication request between two nodes located in different clusters is~1,
and the cost of migrating a node from one cluster to another
is~$\alpha \in \mathbb{Z}^+$.
For any algorithm $ALG$,
its cost,
denoted by $ALG(\sigma)$,
is the total cost of communications and
the cost of migrations on the sequence $\sigma$.

\section{Lower bound} %$\Omega(k\cdot \ell)$ for Competitive Ratio of Any Deterministic Algorithm}

In this section we show a lower bound of $\Omega(k \cdot \ell)$
using an input sequence that allows OPT to move to a perfect partition.
Therefore,
it is a lower bound for OBRP with perfect partition,
which directly implies a lower bound for the (general) OBRP.
The former complements an upper bound in $O(k \cdot \ell)$
for the OBRP with perfect partition which we present in Section \ref{sec:upperbound}.

\begin{theorem}
  The competitive ratio of any deterministic algorithm for OBRP is in $\Omega(k\cdot \ell)$.
\end{theorem}

\begin{proof}
We construct an instance of the problem with $\ell$ clusters 
$\set{ S_1, S_2,\dots , S_{\ell}}, |S_i|  = k$.
Let $I(C)$ denote the cluster where nodes in a component $C$ are located initially.
Let ALG be any online algorithm for OBRP.
We construct the input sequence of requests for ALG as follows.
First,
we issue $k-2$ (internal) requests so that ALG can form a component of $k-1$
nodes on the cluster $S_1$.
Recall that internal requests does not incur ALG any cost.
If ALG at any point splits a component then
we issue  requests between the parts as many as it is necessary for ALG to collocate the parts,
as otherwise ALG is not competitive.
Let $x_0$  be the only single node left on $S_1$ and  $y_0 \in S_2$ be any single node on $S_2$.
Next,
we issue the request between $x_0$ and $y_0$.
Since this is an external request,
ALG joins them into one component and collocates them in some cluster other than $S_1$.
For this,
ALG moves to a new configuration (i.e., partition)
that replaces $x_0$ and $y_0$ with two other single nodes $x_1$ and $y_1$ respectively.
Next,
we issue a request between $x_1$ and the largest component $C$ s.t.~$I(C) = I(x_1)$.
ALG must collocate $x_1$ and $C$ on some cluster other than $S_1$ and
consequently replaces $x_1$ with some other (single) node $x_2$.
We repeat issuing requests between the single node $x_i$ on $S_1$ and the largest component $C'$ s.t.~$I(C')=I(x_i)$,
 until there are only two single nodes left that  originate from the same cluster.
I.e.,
two single nodes $x^*, y^*,I(x^*) = I(y^*)$ s.t.~for any other pair of single nodes
$x'$ and $y'$,
we have $I(x') \neq I(y')$.
At this point there are at most $\ell+1$ single nodes left,
otherwise there would be more pairs of single nodes that were initially in the same cluster.
Given this sequence of requests,
the optimal strategy is to migrate $\set{x_0,y_0}$ to the cluster $I(x^*)$ by
 swapping $\set{x_0,y_0}$ with $\set{x^*,y^*}$.
Hence,
OPT pays $4$ for node migrations and
ALG pays at least one for each node in the sequence $X := x_0, x_1,\dots$.
We exclude at most $(k-1) + ( \ell+1)$ nodes out of $k.\ell$ nodes,
therefore $|X| \geq k.\ell - k - \ell \in \Omega(k\cdot\ell)$.
\end{proof}

\section{Perfect Partition Learner Algorithm}	\label{sec:upperbound}

@@

\begin{algorithm}
    \renewcommand{\algorithmicrequire}{\textbf{Input:}}
    \renewcommand{\algorithmicensure}{\textbf{Output:}}
    \begin{algorithmic}[1]
        \Require 
        $k, \ell$,
        initial partition $P_I$,
        sequence of  requests $\sigma_1, \dots, \sigma_N$ 
        \Ensure A final partition $P_F$ 
        \State For each node $v$ create a singleton component $C_v$ and add it to $\mathcal{C}$
        \State{$P_0 := P_I$}
         \label{line:initcomponents}
        \For {each  request $\sigma_t=\{u,v\}, 1 \leq t \leq N$}
        \State Let $C_1 \ni u$ and $C_2 \ni v$ be the container components
        \If{$C_1 \neq C_2$}
        \State Unite the two components into a single component $C'$ and
        $\mathcal{C} = (\mathcal{C}\setminus\set{C_1, C_2}) \cup ~\set{C'}$ \label{line:mergecomponents}
%        \If{$\mathit{cluster}(C_1, P_{t-1}) \neq \mathit{cluster}(C_2, P_{t-1})$}
        \If{$\mathit{cluster}(C_1, P_{t-1}) \neq \mathit{cluster}(C_2, P_{t-1})$}
        \Comment{if not in the same cluster}
        \State $P_{t} = \mathit{re\mhyphen partition}(P_{t-1}, P_I, \mathcal{C}$) \label{line:rebalance} 
        \Comment{move to $P$ closest to $P_I$}
        \EndIf
        \EndIf
        \EndFor
    \end{algorithmic}
    \caption{Perfect Partition Learner (PPL)}
    \label{alg:ppl}
      \end{algorithm}

@@

Let \emph{GREEDY} be the algorithm that on each inter-cluster request
 moves to a perfect partition closest to the current one.
 That is,
 a partition reachable with the least number of node swaps from the current partition.

@@

\begin{lemma} \label{lemma:k=3}
  The re-balancing cost of GREEDY for $k=3$ is $O(1)$.
  %\maciek{GREEDY = find the cheapest (closest to the current) feasible partition}
  \label{rebalancing-cost}
\end{lemma}

@@

\begin{theorem}
	Algorithm \emph{GREEDY} is $O(\ell)$-competitive when $k=3$.
\end{theorem}

\begin{proof}
	The adversary issues at most $3\ell$ requests, as otherwise the perfect partition would not exist.
	Each request triggers at most one re-balancing in \emph{GREEDY}
	which costs $O(1)$ by Theorem~\ref{rebalancing-cost}.
	Then,
	the total cost of any algorithm is $O(\ell)$.
\end{proof}

We leverage the simple algorithm \emph{GREEDY} for $k=3$
towards an algorithm for the general problem
where the communication graph is not restricted.
Note that in the general model,
an optimal partition is not a necessarily a perfect partition.
We divide the sequence of requests into phases.
A phase terminates when there is no perfect partition of components.
We apply requests to \emph{GREEDY} as before. 
Once the current phase terminates,
we reset all components by removing (forgetting) all revealed edges,
and then we proceed to the next phase.

\begin{theorem} \label{cor:k=3}
	There exists $O(\ell)$-competitive algorithm for the general online problem when $k=3$. 
\end{theorem}
\begin{proof}
	Consider any phase and assume it terminates upon arrival of some request $r$.
	By Lemma  \ref{lemma:k=3},
	\emph{GREEDY} pays $O(\ell)$ during the phase.
	If OPT stays in the same configuration throughout the phase,
	regardless of its partition at the beginning of the phase,        
	it incurs remote communication cost for at least one request $r'$ (possibly $r'=r$) by the end of the phase.
	Else,
	OPT incurs the cost of moving to new partitions.
	In any case,
	OPT pays at least one during the phase and 
	the competitive ratio is $O(\ell)$.
\end{proof}
